---
title: "Photoperiod Notebook"
output: html_notebook
---

#1: Read in required packages and data
```{r}
library(tidyverse)
library(rgeos) #for gCentroid()
library(rworldmap) #for getMap()
library(rworldxtra) #for higher resolution map

pp <- read.csv("./data/appendix1.csv") #read in photoperiod data
pplist <- gsub("_", " ", pp$species) #extract list of study species
rm(pp) #unload photoperiod data from environment

apply_study_scope <- function(df) {
  require(dplyr)
  df %>% 
  filter(species %in% pplist) %>%
  dplyr::filter(is.na(doy) == F) %>% #remove na
  filter(year>=1950) %>% #constrain years
  filter(year<=2017) %>%
  group_by(species, year, location, lat, lon)%>% #for each location*year*species combination,
  slice(which.min(doy))  #take the first day of flowering every year
}
```

NOTE: each source is processed seperately to avoid the "Unknown or uninitialised column" error likely associated with too much in the environment (aka we try to remove unecessary data points as early as possible). If you still get errors running this code, restart R and skip to the the second notebook; where we read straight from the appendix located with the published data ([add data repository link here])


#2: Craine, J.M., Wolkovich, E.M., Gene Towne, E. and Kembel, S.W., 2012. Flowering phenology as a functional trait in a tallgrass prairie. New Phytologist, 193(3), pp.673-682.
```{r}
craine <- read.csv("./data/raw/Craine2012/craine_2012.csv")

craine$location <- "USA"
craine$lat <- as.numeric("39.08") #add location based on paper (coordinates provided)
craine$lon <- as.numeric("-96.56")
craine$latlon_verified <- T #google earth verify approximate location
craine$source <- as.factor("craine") #add source

craine <- apply_study_scope(craine)
flowering_time <- craine
rm(craine)
```

#3: Fitter, A.H. and Fitter, R.S.R., 2002. Rapid changes in flowering time in British plants. Science, 296(5573), pp.1689-1691.
```{r}
fitter <- read.csv("./data/raw/Fitter2002/fitter_2002.csv")

fitter <- gather(fitter, "year", "doy", 2:48)#tidy
fitter$year <- as.integer(str_sub(fitter$year, 2))#remove X from year
colnames(fitter)[1] <- "species" #rename species column
fitter <- fitter %>% filter(is.na(fitter$doy) == F)#remove NAs
fitter$lat <- as.numeric("55.3781") #centre of great britain
fitter$lon <- as.numeric("-3.4360")
fitter$location <- "Britain"
fitter$latlon_verified <- F
fitter$source <- "fitter"#add source

fitter <- apply_study_scope(fitter)
flowering_time <- bind_rows(flowering_time, fitter)
rm(fitter)
```

#4: Park D, Williams A, Law E, Ellison A, Davis C. 2018. Assessing Plant Phenological Patterns in the Eastern United States Over the Last 120 Years. Harvard Forest Data Archive: HF309.
```{r}
park <- read.csv("./data/raw/Harvard_Forest/hf309-01.csv")
park <- park %>% 
  filter(phenophase == "efl") %>% #note that efl = first flower
  transmute(species = gsub("_", " ", name), 
            year = year, 
            doy = doy, 
            location = "USA", 
            lat = as.numeric(county.lat),
            lon = as.numeric(county.lon),
            latlon_verified = F,
            source = "park")

park <- apply_study_scope(park)
flowering_time <- bind_rows(flowering_time, park)
rm(park)
```

#5: Davis C, Ellison A. 2016. Eastern Massachusetts Flowering Phenology 1852-2013. Harvard Forest Data Archive: HF258.

hf258-03:Herbarium
```{r}
davis_h <- read.csv("./data/raw/Harvard_Forest/hf258-03.csv") #herbarium
davis_h <- davis_h %>%
  filter(phenology == "Flowering") %>% #filter flowering only
  transmute(species = genus.species, 
            year = year, 
            doy = doy, 
            lat = NA,
            lon = NA,
            location = locality,
            source = "davis")
```
hf258-04: Field
```{r}
davis_f <- read.csv("./data/raw/Harvard_Forest/hf258-04.csv") #field
davis_f <- davis_f %>%
  filter(phenology == "Flowering") %>% #filter flowering only
  transmute(species = genus.species, 
            year = year, 
            doy = doy, 
            lat = NA,
            lon = NA,
            location = locality,
            source = "davis")
```
Combined
```{r}
davis <- bind_rows(davis_f, davis_h)
rm(davis_f)
rm(davis_h)

#manually specify locations
davis$lat[davis$location == "Middlesex"] <- "42.4672" 
davis$lon[davis$location == "Middlesex"] <- "-71.2874"

davis$lat[davis$location == "Worcester"] <- "42.2626" 
davis$lon[davis$location == "Worcester"] <- "-71.8023"

davis$lat[davis$location == "Barnstable"] <- "41.7003"
davis$lon[davis$location == "Barnstable"] <- "-70.3002"

davis$lat[davis$location == "Essex"] <- "42.6320"
davis$lon[davis$location == "Essex"] <- "-70.7829"

davis$lat[davis$location == "Plymoth"] <- "41.9584"
davis$lon[davis$location == "Plymoth"] <- "-70.6673"

davis$lat[davis$location == "Bristol"] <- "41.7938"
davis$lon[davis$location == "Bristol"] <- "-71.1449"

davis$lat[davis$location == "Norfolk"] <- "42.1195"
davis$lon[davis$location == "Norfolk"] <- "-71.3251"

davis$lat[davis$location == "Suffolk"] <- "42.3523"
davis$lon[davis$location == "Suffolk"] <- "-71.0504"

davis$lat[davis$location == "Berkshire"] <- "42.3118"
davis$lon[davis$location == "Berkshire"] <- "-73.1822"

davis$lat[davis$location == "Dukes"] <- "41.4040"
davis$lon[davis$location == "Dukes"] <- "-70.6693"

davis$lat[davis$location == "Hampshire"] <- "42.3471"
davis$lon[davis$location == "Hampshire"] <- "-72.6624"

davis$lat[davis$location == "Franklin"] <- "42.0834"
davis$lon[davis$location == "Franklin"] <- "-71.3967"

davis$lat[davis$location == "Hampden"] <- "42.0640"
davis$lon[davis$location == "Hampden"] <- "-72.4134"

davis$lat[davis$location == "Nantucket"] <- "41.2835"
davis$lon[davis$location == "Nantucket"] <- "-70.0995"

davis <- apply_study_scope(davis)
davis$lat <- as.integer(davis$lat) #change to integer
davis$lon <- as.integer(davis$lon)
flowering_time <- bind_rows(flowering_time, davis)
rm(davis)
```

#6: Abu-Asab, M.S., Peterson, P.M., Shetler, S.G. and Orli, S.S., 2001. Earlier plant flowering in spring as a response to global warming in the Washington, DC, area. Biodiversity & Conservation, 10(4), pp.597-612.
```{r}
abuasab <- read.csv("./data/raw/AbuAsab2001/abuasab_2001_doy.csv")
abuasab <- gather(abuasab, "year", "doy", 2:30)

abuasab <- abuasab %>%
  transmute(species = Species, 
            year = as.integer(gsub("X", "", year)),
            doy = doy, 
            location = "Washington DC", 
            lat = 38.9072, #set location as washington dc
            lon = -77.0369,
            latlon_verified = F,
            source = "abuasab")

abuasab <- apply_study_scope(abuasab)
flowering_time <- bind_rows(flowering_time, abuasab)
rm(abuasab)

```

#7:Rocky Mountain Biological Lab: Long-term flowering phenology and abundance data at Gothic, Colorado
DOI 10.17605/OSF.IO/JT4N5
```{r}
rmbl <- read.csv("./data/raw/RMBL Phenology/RMBL_Phenology_1974-2017_Zeng.csv")

rmbl$lat <- "38.9619215" #location is centred on plot RM1 since all plots are fairly close to eachother and so would contain the same population
rmbl$lon <- "-106.9865916"

rmbl <- rmbl %>%
    transmute(species = species, 
            year = year,
            doy = doy, 
            location = plot, 
            lat = as.integer(lat),
            lon = as.integer(lon),
            latlon_verified = F,
            source = "rmbl")

rmbl <- apply_study_scope(rmbl)
flowering_time <- bind_rows(flowering_time, rmbl)
rm(rmbl)
```

#8: [UK Nature's Calender Woodland Trust]
```{r}
uknc <- bind_rows(
  read.csv("./data/raw/UKNCWT/uknc_1950-1999.csv"),
  read.csv("./data/raw/UKNCWT/uknc_2000-2006.csv"),
  read.csv("./data/raw/UKNCWT/uknc_2007-2016.csv")
  )

#unique(uknc$SpeciesLatinName) #add missing latin names
uknc$SpeciesLatinName[uknc$SpeciesName == "Plum"] <- "Prunus domestica"
uknc$SpeciesLatinName[uknc$SpeciesName == "Pignut, Earthnut"] <- "Conopodium majus"
uknc$SpeciesLatinName[uknc$SpeciesName == "Wild Cherry, Gean"] <- "Prunus avium"
uknc$SpeciesLatinName[uknc$SpeciesName == "Wood Avens, Herb Robert"] <- "Geum urbanum"
uknc$SpeciesLatinName[uknc$SpeciesName == "Heather, Ling"] <- "Calluna vulgaris"
uknc$SpeciesLatinName[uknc$SpeciesName == "Turnip, Naven"] <- "Brassica rapa subsp rapa"

uknc1 <- uknc %>% #group 1: has latitude
  filter(is.na(Latitude) == F)

uknc2 <- uknc %>%  #group 2: has no latitude but eastings and northings
  filter(is.na(Latitude) == T) %>% 
  filter(is.na(Easting) == F)

ukgrid = "+init=epsg:27700" #define the current grid system (OSGB1936 / British National Grid - United Kingdom Ordnance Survey)
latlon = "+init=epsg:4326" #define WGS84, the grid sytem that we are using

library(raster)
library(rgdal)

uknc2.spatial <- SpatialPoints(data.frame(x = uknc2$Easting, y = uknc2$Northing))#create spatial points

proj4string(uknc2.spatial) <- CRS(ukgrid) #add crs (it was auto-detected but still best practice)

uknc2.spatial <- spTransform(uknc2.spatial, CRS(latlon))#reproject

uknc2$Longitude <- uknc2.spatial@coords[,1]
uknc2$Latitude <- uknc2.spatial@coords[,2]
  
rm(uknc2.spatial)

uknc3 <- uknc %>% #group 3: has no latitude and no eastings and northings
    filter(is.na(Latitude) == T) %>% 
  filter(Easting == 0)

uknc3$Latitude <- 53.8337 #set location as the middle of the UK
uknc3$Longitude <- 2.4290

uknc <- bind_rows(uknc1, uknc2, uknc3) #recombine
rm(uknc1)
rm(uknc2)
rm(uknc3)


uknc <- uknc %>%
    transmute(species = SpeciesLatinName, 
            year = ObservationYear,
            doy = ObservationDay, 
            location = loccounty, 
            lat = Latitude,
            lon = Longitude,
            latlon_verified = F,
            source = "uknc")

uknc <- apply_study_scope(uknc)#apply study scope
flowering_time <- bind_rows(flowering_time, uknc)
rm(uknc)

```


#9: [US Phenology Network]
```{r}
uspn <- read.csv("./data/raw/US Phenology Network/status_intensity_observation_data.csv")

uspn <- uspn %>% 
  filter(Phenophase_ID == "77"|Phenophase_ID == "412"|Phenophase_ID == "428")#take first bloom for each species
  
uspn <- uspn %>%
    transmute(species = paste(Genus, Species), #combine genus species
            year = as.integer(str_sub(Observation_Date, start = -4)), #extract year by taking last 4 characters of the date
            doy = Day_of_Year, 
            location = State, 
            lat = Latitude,
            lon = Longitude,
            latlon_verified = F,
            source = "uspn")

uspn <- apply_study_scope(uspn)
flowering_time <- bind_rows(flowering_time, uspn)
rm(uspn)

```


#10: [PEP 725 reference] We have only identified location to country, but more detail can be found by matching PEP725 station codes (500 IPG records (before filters) were removed this way)
```{r}
#comment code is what was used to build the useful dataset from raw, it is saved so there is no need to rerun normally

#pep725.list <- list.files(path = "./data/raw/EU_PEP725/PEP725_FT/", full.names = F)#create list of csvs


#add_filename <- function(arg_filename, input_folder = "./data/", output_folder = "./output/"){
#arg_data <-read.csv2(paste0(input_folder, arg_filename))#read file, note that files are in quropean format where semicolon is used instead of comma
#
#arg_data <- mutate(arg_data, filename = arg_filename)#add new column
#
#  write.csv(arg_data, paste0(output_folder,"edited_", arg_filename)) #write new file in output folder with new name
#}

#add_filename("PEP725_AT_Beta_vulgaris.csv", "./data/raw/EU_PEP725/PEP725_FT/", "./data/output/") #test on the first one

#pep725.list %>%
#  map(add_filename, input_folder = "./data/raw/EU_PEP725/PEP725_FT/", output_folder = "./data/output/pep725/") #apply function to everything

pep725 <- list.files("./data/output/pep725/", full.names = T) %>% map_df(read.csv)#read in the whole folder

pep725 <- pep725 %>% filter(BBCH == "60") #filter for first flowering (BBCH = 60)

pep725$location <- str_sub(pep725$filename, start = 8, end = 9) # obtain country code
pep725$species <- str_sub(pep725$filename, start = 11, end = -5) #species name

pep725.countrycodes <- read.csv("./data/2DigitCountryCodes_ISO_3166_1.txt") #load in country codes from https://datahub.io/core/country-list#resource-data
colnames(pep725.countrycodes) <- c("country", "location")#add country based on filename
pep725 <- right_join(pep725.countrycodes, pep725, by = "location")
rm(pep725.countrycodes)
#pep725$location[is.na(pep725$country) == T] #identify missing countries
pep725$country[pep725$location == "UK"] <- "United Kingdom"
pep725 <- pep725 %>% filter(location != "IP") #IP = IPG, the international phenological gardens network, that includes many different locations

wmap <- getMap(resolution="high") #get world map
countrycoords <- gCentroid(wmap, byid=TRUE) #get centroids of countries
countrycoords <- as.data.frame(countrycoords)
countrycoords <- rownames_to_column(countrycoords, var = "country")#turn the row names into a proper column

pep725 <- left_join(pep725, countrycoords, by = "country")#add coordinates to country names
pep725$x[pep725$country == "Macedonia, the Former Yugoslav Republic of"] <- "41.195558"#manually change the coordinates of places that aren't on the list
pep725$y[pep725$country == "Macedonia, the Former Yugoslav Republic of"] <- "22.715024"

rm(wmap)
rm(countrycoords)

pep725$species <- gsub("_", " ", pep725$species) #change species format

pep725$species[pep725$species == "Betula(Betula pendula (B. verrucosa  B. alba))"] <- "Betula pendula" #sub species names without extras
pep725$species[pep725$species == "Fagus(Fagus sylvatica)"] <- "Fagus sylvatica"
pep725$species[pep725$species == "Pinus(Pinus sylvestris)"] <- "Pinus sylvestris"
pep725$species[pep725$species == "Betula(Betula pubescens)"] <- "Betula pubescens"
pep725$species[pep725$species == "Punica granatum(Punica granatum Barski)"] <- "Punica granatum"

pep725 <- pep725 %>%
    transmute(species = species, #combine genus species
            year = YEAR, #extract year by taking last 4 characters of the date
            doy = DAY, 
            location = country, 
            lat = as.numeric(y),
            lon = as.numeric(x),
            latlon_verified = F,
            source = "pep725")

pep725 <- apply_study_scope(pep725)
flowering_time <- bind_rows(flowering_time, pep725)
rm(pep725)

```

#Save final dataset for easy access
```{r}
write.csv(flowering_time,"./data/appendix2.csv", row.names = F)
rm(apply_study_scope)
rm(pplist)
rm(flowering_time)
```
