---
title: "PhotoperiodNotebook5_PhylogeneticAnalysis"
output: html_notebook
---
This code aims to repeat the analysis of flowering time change and photoperiod, with added phylogenetic information (since we found a phylogenetic signal in notebook 4. 

If df is not loading in, run the start of photoperiod notebook 2 to generate the combined dataframe.

In order to include this phylogenetic information, we have had to switch to a per-species analysis instead of the per-photoperiod-group analysis performed previously. This change unfortunately reduced the amount of data we could use in the analysis but will allow us to draw new conclusions as to whether photoperiod sensitivity acts independently from phylogeny, since it is such a plastic trait.

For more information, please contact:
karen.zeng@unsw.edu.au

##load in packages
```{r warning=FALSE}
library(tidyverse) #dplyr for general manipulation
library(lme4) #for general linear model stuff
library(nlme) #mixed effect models
library(broom) #for do function

library(ggplot2) #graphs :)

library(ape) #general phylogenetic manipulation and tree
library(phytools) #phylANOVA for anova with phylogenetic component

```

##load in datasets
```{r}
pp <- read.csv("./data/appendix1.csv") #read in combined photoperiod data

pp <- pp %>% 
  group_by(species)%>% #for each species
  slice(which.min(order))  #find most accurate photoperiod record

ft <- read.csv("./data/appendix2.csv") #read completed flowering time data

df <- left_join(ft, pp, by = "species") #join the two datasets

temp <- df %>% group_by(species) %>% count()

pp <- left_join(pp, temp, by = "species")#add number of flowering time records available per species

pp$simple_photoperiodism <- "other" #if something is not ld, sd or na it will be grouped as 'other'
pp$simple_photoperiodism[pp$photoperiodism == "na"] <-"na"
pp$simple_photoperiodism[pp$photoperiodism == "sd"] <-"sd"
pp$simple_photoperiodism[pp$photoperiodism == "ld"] <-"ld"

pp <- pp[pp$simple_photoperiodism != "other",] #filter out 'other' group

rm(temp)#remove temporary datframe

```


```{r}
species_summary <- df %>% 
  group_by(species) %>% #for each species repeat analysis
  do({ summ <- summary(lm(doy ~ year + lat, .)) #make linear model and take a summary
       summarize(.,intercept = summ[["coefficients"]][1], #take only the numbers we are interested in from each lm
                 intercept_p = summ[["coefficients"]][10], #ie what are the effects of each variable
                 year_effect = summ[["coefficients"]][2],  #and how certain we are of each (p-val)
                 year_p = summ[["coefficients"]][11],
                 lat_effect = summ[["coefficients"]][3],
                 lat_p = summ[["coefficients"]][12])}) %>% ungroup() #ungroup to put it all in one table


species_summary <- left_join(species_summary, pp[, c(1, 2, 7, 9)], by = "species") #put the photoperiod info back in
species_summary$significant_yn <- species_summary$year_effect < 0.05
```

ggplot
```{r}
ggplot(data = species_summary, aes(x = simple_photoperiodism, y = year_effect)) + geom_point(aes(alpha = 0.3, colour = year_effect < 0.05))
```

#summary
Results will greatly change based on n weighting method but I think log is reasonable.
A lot of data is lost in this analysis
But we can correct for phylogeny at last.

That year effect = -6 datapoint is sunflowers. It only has 5 points so a bit borderline but the data looks reasonable.

Can look into in further sometime?

#Prepare tree things
```{r}
tree <- read.tree("./data/ALLMB.txt") #read in tree

species_summary$species <- gsub(" ", "_", species_summary$species)#change species name to fit phylo trees

tree <- drop.tip(tree, tree$tip.label[-na.omit(match(species_summary$species, tree$tip.label))]) #trim tree to our species for ease of use

species_summary$matchtree <- match(species_summary$species, tree$tip.label)
species_summary<- species_summary[is.na(species_summary$matchtree) == F,] #trim species list to tree

row.names(species_summary) <- species_summary$species #set row names as species so namecheck works
namecheck <- name.check(phy = tree, data = species_summary) #check between tree and rownames of species summary which should be the species names now 

#make it so order of dataframe matches order of tree
species_summary <- species_summary%>%arrange(match(species_summary$species, tree$tip.label))
```


#Make a new dataframe and tree with all regression lines and use it to trim a new tree
```{r}
species_summary_finite <- as.data.frame(species_summary[is.finite(species_summary$year_effect),]) #remove non-finite results (na, nan and NULL) and turn it from a tibble to a dataframe

tree_finite <- drop.tip(tree, tree$tip.label[-na.omit(match(species_summary_finite$species, tree$tip.label))]) #trim tree again

tree_finite$node.label[tree_finite$node.label == ""] <- which(tree_finite$node.label == "") #label unnamed nodes based on node number

row.names(species_summary_finite)<- species_summary_finite$species #makes species names the row names again for combining the dataframes
```

##Perform ANOVA between simple photoperiodism groups with all the finite regression lines 
```{r}
simple_photoperiodism_vector <- species_summary_finite$simple_photoperiodism
year_effect_vector <- species_summary_finite$year_effect

aov_species_summary_finite <- phylANOVA(tree = tree_finite, x = simple_photoperiodism_vector, y = year_effect_vector, nsim = 9999, posthoc = TRUE)
```

#Perform comparison ANOVA without phylo information
```{r}
aov_species_summary_NOPHYLO <- aov(year_effect ~ simple_photoperiodism, data = species_summary_finite)
```

#Compare the anovas
```{r}
aov_species_summary_finite

summary(aov_species_summary_NOPHYLO)
```

#Results and Graphs

From the comparison of the ANOVAs above, the difference between groups is no longer significant.

This muddies the water a bit, but it makes sense that large scale trends in photoperiod sensitivity are phylogenetically conserved, and some clades tend to have the same daylength cues for flowering. After all, the evolution of flowering cue mechanisms is constantly happening, and is based on a whole load of interlocking parts working together and apart. 

-----------------------------------------------------------------




