---
title: "R Notebook"
output: html_notebook
---

###PHOTOPERIOD SENSITIVITY AND TRAITS

#load photoperiod sensitivity data
```{r}
library(tidyverse)
library(ggplot2) #for graphs
library(viridis) #for colours

viridis(12, direction = -1) #create palette

pp <- read.csv("./data/appendix1.csv") #read in photoperiod data

pp <- pp %>% 
  group_by(species)%>% #for each species
  slice(which.min(order))  #find most accurate photoperiod record
```

#load woodiness and longevity data from [GlobalWoodinessDatabase] and [TRY]
```{r}
wood <- read.csv("./data/GlobalWoodinessDatabase.csv")

wood <- transmute(wood, species = gs, woodiness = woodiness)

wood$woodiness[wood$woodiness == "H"] <- "herbaceous"
wood$woodiness[wood$woodiness == "W"] <- "woody"

table(wood$woodiness)

longevity <- read.csv(file = "./data/TRYlongevity.csv")

longevity <- longevity %>% 
  mutate(ann_per = case_when(OrigValueStr == "annual" | 
                               OrigValueStr == "Annual" | 
                               OrigValueStr == "annuals" |
                               OrigValueStr == "summer annuals" |
                               OrigValueStr == "always annual" |
                               OrigValueStr == "winter annuals" |
                               OrigValueStr == "annual-winter annual" |
                               OrigValueStr == "winter annual" |
                               (OriglName == "Life history" & OrigValueStr == "1" ) |
                               (OriglName == "Plant phenology: Annual" & OrigValueStr == "yes" )     ~   "annual",  ######annuals
                             OrigValueStr == "perennial" | 
                               OrigValueStr == "Perennial" | 
                               OrigValueStr == "perennials" |                                   
                               OrigValueStr == "always pluriennial-pollakanthic" | 
                               (OriglName == "Plant phenology: Biennial" & OrigValueStr == "yes" ) | 
                               OrigValueStr == "perennial < 20 years" | 
                               OrigValueStr == "woody" | 
                               OrigValueStr == "perennial/woody" | 
                               OrigValueStr == "perennial > 20 years" | 
                               OrigValueStr == "poly-annuals > 50 years (long-lived perennials)" | 
                               OrigValueStr == "always biennial, always pluriennial-hapaxanthic" | 
                               OrigValueStr == "always biennial, always pluriennial-pollakanthic" | 
                               OrigValueStr == "tree" | 
                               OrigValueStr == "shrub" | 
                               OrigValueStr == "always pluriennial-hapaxanthic, always pluriennial-pollakanthic" | 
                               OrigValueStr == "always pluriennial-hapaxanthic" | 
                               OrigValueStr == "biennial" | 
                               OrigValueStr == "annual/biennial" | 
                               OrigValueStr == "poly-annuals < 5 years (short-lived perennials)" | 
                               OrigValueStr == "Biennial" | 
                               OrigValueStr == "biennial/perennial" | 
                               OrigValueStr == "always biennial" | 
                               OrigValueStr == "biennial-perennial" | 
                               OrigValueStr == "sometimes biennial, always pluriennial-hapaxanthic, sometimes pluriennial-pollakanthic" | 
                               OrigValueStr == "sometimes biennial, sometimes pluriennial-hapaxanthic, always pluriennial-pollakanthic" | 
                               OrigValueStr == "biennial/perennial/woody" | 
                               OrigValueStr == "sometimes biennial, always pluriennial-pollakanthic" | 
                               OrigValueStr == "poly-annuals 5-50 years (medium-lived perennials)" | 
                               (OriglName == "Plant phenology: Perennial" & OrigValueStr == "yes" )| 
                               (OriglName == "Plant phenology: Annual" & OrigValueStr == "no" )    ~   "perennial"  ###perennials
  ))

longevity <- transmute(longevity, species = SpeciesName, life_history = ann_per) #isolate info

longevity <- na.omit(longevity) #remove NAs

longevity <- longevity[!duplicated(longevity$species), ] #remove duplicates

longevity$simple_photoperiodism <- factor(longevity$simple_photoperiodism, levels = c("sd", "ld", "na")) #relevel factors for consistency

```


#combine the datasets
```{r}
pp <- left_join(pp, wood, by = "species") #attach woodiness dataset to main 
pp <- left_join(pp, longevity, by = "species") #attach longevity dataset to main

rm(longevity) #remove both datasets from environment
rm(wood)
```



#Life History Analysis
```{r}
longevity <- pp %>% #Tally life histories
  filter(!is.na(life_history))%>% #remove records without life history
  filter(simple_photoperiodism != "other") %>% #remove records with unclear photoperiodism
  group_by(simple_photoperiodism, life_history) %>%
  tally()

longevity_chisq <- chisq.test(longevity$n) #Perform chi-square test

perennial_pie <- longevity %>% #choose our dataframe
  filter(life_history == "perennial") %>% #Filter so we are only looking perennials
  ggplot(aes(x = "", y = n, fill = simple_photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Perennial")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= mypalette[11], "ld"= mypalette[8], "na"= mypalette[3])) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels

annual_pie <- longevity %>% #choose our dataframe
  filter(life_history == "annual") %>% #Filter so we are only looking at annuals
  ggplot(aes(x = "", y = n, fill = simple_photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Annual")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= mypalette[11], "ld"= mypalette[8], "na"= mypalette[3])) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels
```

Look at results
```{r}
longevity_chisq
longevity_chisq$residuals
longevity_chisq$expected
longevity_chisq$observed

perennial_pie
annual_pie

```

Annual species seem to be more likely to be photoperiod-sensitive overall, and seem to be more short day than long day, although they exist in both groups.

#Woodiness analysis
```{r}
woodiness <- pp %>% #Tally woodiness
  filter(!is.na(woodiness))%>% #remove records without woodiness
  filter(woodiness != "variable") %>% #remove records with unclear woodiness
  filter(simple_photoperiodism != "other") %>% #remove records with unclear photoperiodism
  group_by(simple_photoperiodism, woodiness) %>%
  tally()

woodiness_chisq <- chisq.test(woodiness$n) #Perform chi-square test

woody_pie <- woodiness %>% #choose our dataframe
  filter(woodiness == "woody") %>% #Filter so we are only looking at herbaceous species
  ggplot(aes(x = "", y = n, fill = simple_photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Woody")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= mypalette[11], "ld"= mypalette[8], "na"= mypalette[3])) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels


herbaceous_pie <- woodiness %>% #choose our dataframe
  filter(woodiness == "herbaceous") %>% #Filter so we are only looking at this type of plant
  ggplot(aes(x = "", y = n, fill = simple_photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Herbaceous")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= mypalette[11], "ld"= mypalette[8], "na"= mypalette[3])) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels

```

Look at results
```{r}
woodiness_chisq
woodiness_chisq$residuals
woodiness_chisq$expected
woodiness_chisq$observed

woody_pie
herbaceous_pie
```
Both herbaceous and woody species are just as likely to be photoperiod sensitive, but it seems that woody species are more likely to be short day, and conversely herbaceous species are more likely to be long-day.

### PHYLOGENETIC SIGNAL ANALYSIS

#load in photoperiod sensitivity data again
```{r}
library(viridis)#colour palette
library(tidyverse)#general data manipulation
library(ggplot2) #for graphs
library(ggtree) #ggplot for phylogenetic trees

library(ape) #read.tree(), ditomulti() and manipulating phylogenetic trees
library(phylobase) #phylo4d objects

#library(phytools)#phylosig()
#library(caper) #phylo.d()
#library(phylosignal)#phylogenetic signal

library(picante)#phylogenetic signal for nonbinary traits


mypalette <- viridis(12, direction = -1)

pp <- read.csv("./data/appendix1.csv") #read in photoperiod data

#remove other

pp <- pp %>% 
  group_by(species)%>% #for each species
  subset(simple_photoperiodism != "other")%>% #remove records with complex photoperiod sensitivity
  slice(which.min(order))  #find most accurate photoperiod record

pp$species <- gsub(" ", "_", pp$species) #underscore is preferred over space when working with trees
```

#load in phylogenetic tree from smith and brown (2018) (https://doi.org/10.1002/ajb2.1019)
```{r}
  tree <- read.tree("./data/ALLMB.txt")

  species_list <- pp$species #create list of species
  
  tree <- drop.tip(tree, tree$tip.label[-na.omit(match(species_list, tree$tip.label))]) # Prune tree by dropping all tips (species) that are not on our species list
   
  tree_order <- tree$tip.label #Sort the species data into the same order as the tree for graphing later on
  pp_sorted <- data.frame(Reduce(rbind, tree_order)) #try to combine these two lines 
  names(pp_sorted)[1] <- "species" #change column name
  pp_sorted <- left_join(pp_sorted, pp, by = "species") #attach data to new order
   
  rm(pp) #remove unused parts
  rm(species_list)
  rm(tree_order)

#    tree2=di2multi(tree) #fix multichotomies
#    tree <- makeLabel(tree) #fix node error 
    
  tree4d <- phylo4d(tree, (pp_sorted$simple_photoperiodism))  
       
```

#Fixing up the data: 
I think I found a known error with clade names where un-named nodes are treated as having duplicated names.
The fix was to numerically name each node so it has a unique name
Workaround was found on: http://grokbase.com/t/r/r-sig-phylo/11ce1b8stn/doubts-and-problems-bug-when-using-comparative-data-and-pgls-functions-in-caper%E2%80%8F

# binary photoperiod sensitivity phylogenetic signal (unused) that doesn't woprk for multiple levels of photoperiodism
```{r}
#tree4d_bin <- phylo4d(tree, (pp_sorted$sensitivity_yn))
  
#phylo.d(pp_sorted, tree, names.col = species, binvar = sensitivity_yn)#d statistic  

#local.i <- lipaMoran(tree4d, prox.phylo = "nNodes", as.p4d = TRUE) #local phylogenetic signal
#points.col <- lipaMoran(tree4d, prox.phylo ="nNodes")$p.value #EXTRACT PVAL
#points.col <- ifelse(points.col < 0.05, "red", "black") #SIGNIFicant p val limit is 0.05

#dotplot.phylo4d(local.i, dot.col = points.col)
  
```

This analysis was not used in the final product after switching to a more detailed look at photoperiodism

The below analysis was used instead.

First, to prepare the tree for analysis by setting a minimum branch length of 0.001 so the tree is dichotomous and doesn't have multichotomies. We will also repeat analysis where instead of forcing a positive value to zero branch lengths we will drop them entirely (tree2)

```{r}
tree <- multi2di(tree) #make tree dichotomous

tree$edge.length[tree$edge.length == 0] <- 0.001 #set edge lengths of zero to a number lower than the minimum

tree2 <- tree #option to repeat analysis with multichomies dropped instead

df_temp <- as.data.frame(tree2$edge)
df_temp$edge_length <- tree2$edge.length
multichotomy_nodes <-  df_temp$V1[df_temp$edge_length == 0.001]

drop_list <- c("Calibrachoa_heterophylla") #start list

for (i in multichotomy_nodes){
  temp <- extract.clade(tree2, node = i)
  drop_list <- c(drop_list, temp$tip.label)
} #fill list of tips associated with nodes

drop_list <- unique(drop_list) #make sure there aren't duplicates

tree2 <- drop.tip(tree2, drop_list) #finally drop all tips associated with zero length edges

pp_sorted2 <- pp_sorted[!(pp_sorted$species %in% drop_list),] #do the same for photoperiod data

rm(df_temp)#cleanup
rm(temp)
rm(drop_list)
rm(multichotomy_nodes)
rm(i)
```

Next, to see if there is a signal

Please download and add the entire folder for code for delta statistic into the R folder for this project before continuing to the next step. 

Code for delta statistic from https://github.com/mrborges23/delta_statistic with methods adapted from associated publication at https://doi.org/10.1093/bioinformatics/bty800.

Since the examples provided in the paper did not go up to our sample size (477), I generated my own comparison data. I compared the delta from our dataset to dummy datasets with phylogenetic signal partially or fully randomised.

```{r}
source("./R/delta_statistic-master/code.R")

#SOME PARAMETERS from the example
#lambda0 <- 0.1   #rate parameter of the proposal 
#se      <- 0.5   #standard deviation of the proposal
#sim     <- 10000 #number of iterations
#thin    <- 10    #we kept only each 10th iterate 
#burn    <- 100   #100 iterates are burned-in

delta_stat <- delta(trait = pp_sorted$simple_photoperiodism, #calculate delta
                    lambda0 = 0.1,
                    se = 0.5,
                    sim = 10000,
                    thin = 10,
                    burn = 100,
                    tree = tree
                    )

weakest_trait <- sample(pp_sorted$simple_photoperiodism)

delta_weakest <- delta(trait = weakest_trait, #calculate delta but randomise all
                    lambda0 = 0.1,
                    se = 0.5,
                    sim = 10000,
                    thin = 10,
                    burn = 100,
                    tree = tree)


weaker_trait <- pp_sorted$simple_photoperiodism # make copy of data
random_half <- sample(1:477)[1:238]#randomly select half the rows
for (i in random_half) {
 weaker_trait[i] <- weakest_trait[i]
}#replace the selected half with the randomised trait values
rm(random_half)

delta_weaker <- delta(trait = weaker_trait, #calculate delta but randomise half
                    lambda0 = 0.1,
                    se = 0.5,
                    sim = 10000,
                    thin = 10,
                    burn = 100,
                    tree = tree)

delta_stat 
delta_weaker
delta_weakest

barplot(c(delta_stat, delta_weaker, delta_weakest), names.arg=c("Photoperiod", "Half Randomised", "Full Randomised"), ylab = "delta statistic", xlab = " Phylogenetic Signal Tests")
```
The delta statistic of photoperiod sensitivity as a three leveled trait is a lot higher than a model


#repeat of analysis but with tree2
```{r}
delta_stat2 <- delta(trait = pp_sorted2$simple_photoperiodism, #calculate delta
                    lambda0 = 0.1,
                    se = 0.5,
                    sim = 10000,
                    thin = 10,
                    burn = 100,
                    tree = tree2
                    )

weakest_trait2 <- sample(pp_sorted2$simple_photoperiodism)

delta_weakest2 <- delta(trait = weakest_trait2, #calculate delta but randomise all
                    lambda0 = 0.1,
                    se = 0.5,
                    sim = 10000,
                    thin = 10,
                    burn = 100,
                    tree = tree2)


weaker_trait2 <- pp_sorted2$simple_photoperiodism # make copy of data
random_half2 <- sample(1:344)[1:172]#randomly select half the rows
for (i in random_half2) {
 weaker_trait2[i] <- weakest_trait2[i]
}#replace the selected half with the randomised trait values
rm(random_half2)

delta_weaker2 <- delta(trait = weaker_trait2, #calculate delta but randomise half
                    lambda0 = 0.1,
                    se = 0.5,
                    sim = 10000,
                    thin = 10,
                    burn = 100,
                    tree = tree2)

delta_stat2
delta_weaker2
delta_weakest2

barplot(c(delta_stat2, delta_weaker2, delta_weakest2), names.arg=c("Photoperiod", "Half Randomised", "Full Randomised"), ylab = "delta statistic", xlab = " Multichotomies Dropped Comparison")
```
The numbers are a little different, but the result is the same, the delta statistic of photoperiod sensitivity is much higher than that of a trait with no phylogenetic signal.

##Things i tried that didn't work as well
```{r}
phylosig(tree, pp_sorted$photoperiodism, method = "lambda") #doesn't work

for (i in unique(pp_sorted_simplified$simple_photoperiodism)){
  pp_sorted_simplified[,paste0(i)]=ifelse(pp_sorted_simplified$simple_photoperiodism==i,1,0)
}
rm(i) #for each type of photoperiodism, make a new column of presence/absence

multiPhylosignal(x = pp_sorted_mps$simple_photoperiodism, phy = tree2, reps = 50)

picante::phylosignal(pp_sorted_mps$simple_photoperiodism, tree2, reps = 50, checkdata = F)

pp_sorted_mps <- pp_sorted[, c(1, 7)]


pp_sorted_mps <- pp_sorted_mps[, -2] #drop simple_photoperiodism

multiPhylosignal(x = pp_sorted_mps, phy = tree, reps = 50)

testtree4d <- phylo4d(tree, (pp_sorted_simplified)) 

```


#prep to make graph
```{r}
tree4d@data[["dt"]] <- factor(tree4d@data[["dt"]], levels = c("sd", "ld", "na")) #relevel data for graphs

tree_plot_rbase <- ggtree(tree4d)+
  geom_tiplab(size = 1.8, offset = -0.9)
```

#graph with node labels to help with labels
```{r}
tree_plot_nodes <- tree_plot_rbase +
  scale_fill_discrete(name="Photoperiod sensitivity") +
  geom_tippoint(aes(color=dt), size=1.5, alpha=0.5, show.legend = NA)+
  scale_color_manual(name = "Photoperiod sensitivity", 
                     values = c(mypalette[11], mypalette[8], mypalette[3]))+
  theme(legend.position = "none") + 
  geom_text(aes(label=node), size = 2, hjust=-.3)
```
Print this in 20*40 to actually be able to read the nodes!

#final colour key version of graph
```{r}
tree_plot_rectangular <- tree_plot_rbase +
  scale_fill_discrete(name="Photoperiod sensitivity") +
  geom_tippoint(aes(color=dt), size=1.5, alpha= 1, show.legend = NA)+
  scale_color_manual(name = "Photoperiod sensitivity", 
                     values = c(mypalette[11], mypalette[8], mypalette[3]))+
  theme(legend.position = "none") + #labels
    geom_strip(taxa1 = 199, taxa2 = 200, 
               barsize = 0.8, offset = 14,
               color = mypalette[8],
               label = " Calibrochoa")+ 
    geom_strip(taxa1 = 40, taxa2 = 64, 
               barsize = 0.8, offset = 22,
               color = "gray",
               label = " Faboideae")+ 
    geom_strip(taxa1 = 40, taxa2 = 47, 
               barsize = 0.8, offset = 14,
               color = mypalette[8],
               label = " IRLC")+ 
    geom_strip(taxa1 = 56, taxa2 = 64, 
               barsize = 0.8, offset = 14,
               color = mypalette[11],
               label = " MILL")+ 
    geom_strip(taxa1 = 24, taxa2 = 25, 
               barsize = 0.8, offset = 14,
               color = mypalette[3],
               label = " Geum")+ 
    geom_strip(taxa1 = 399, taxa2 = 435, 
               barsize = 0.8, offset = 14,
               color = "gray",
               label = " Poöideae")

tree_plot_rectangular
```

#make circular graph as well for fun
```{r}
tree_plot_cbase <- ggtree(tree4d, layout = "circular")+#make circular tree
    geom_tiplab2(size = 1.5, offset = 1)#add species names

tree_plot_circle <- tree_plot_cbase +
  geom_tippoint(aes(color=dt), size=1.5, alpha=.75, show.legend = NA)+
  scale_color_manual(values = c(mypalette[11], mypalette[8], mypalette[3]))+
  scale_fill_discrete(name="Photoperiod sensitivity",
                         labels=c("Short day", "Long day", "Day neutral"))+
  theme(legend.position = "none")
  
tree_plot_cbase
tree_plot_circle
```

