---
title: "R Notebook"
output: html_notebook
---

#load data
Photoperiod data:
- species
- photoperiodism: photoperiod sensitivity requirements(ad, id, ld, sd, ldsd, lsd, sd, sld, yes (exists in unknown form), na)
- cultivar: if present, cultivar name
- reference: source of information
- notes: additional notes on photoperiod sensitivity
- sensitivity_yn: binary variable of whether photoperiod sensitivity in any form exists in species
- order: order in which records are considered, non cultivars and more specific records are prioritised

Flowering time data:
- species
- year: year of record
- doy: day of year
- source: source of information

```{r}
library(tidyverse)
library(ggplot2) #for graphs

library(viridis)
mypalette = viridis(12) #viridis over 12 segments, 1=yellow 4=green 8=blue 12=purple

pp <- read.csv("./Data/appendix1.csv") #read in photoperiod data

pp <- pp %>% 
  group_by(species)%>% #for each species
  slice(which.min(order))  #find most accurate photoperiod record

ft <- read.csv("./data/appendix2.csv") #read completed flowering time data

df <- left_join(ft, pp, by = "species") #join the two datasets
rm(pp)
rm(ft)

```
#Remove unused records and relevel factors so day neutral is the base.

```{r}
df <- df %>% subset(simple_photoperiodism != "other")#remove data without long, short or no daylenth requirement. 

df$simple_photoperiodism <- as.factor(df$simple_photoperiodism) #make it a factor
df$simple_photoperiodism <- relevel(df$simple_photoperiodism, "na") #relevel so na is the reference factor
```

#a quick summary and map of our data
```{r}
summary(df_final)

library(sf) #for applying graphs to spatial data (simple version of sp)
library(rgeos) #for gCentroid()
library(rworldmap) #for getMap()
library(rworldxtra) #for higher resolution map

map_base <- ggplot() + #specify map base
  borders("world", colour="gray50", fill="gray50")

map_pp <- map_base + #use map base
  geom_jitter(data = df, aes(x = lon, y = lat, colour = simple_photoperiodism), size = 0.05, alpha = 0.1, shape = 1, width = 0.4, height = 0.15) +#layer points on map
  xlim(-125, 45)+ #set x limits based on summary
  ylim(20, 65)+#set y limits based on summary
  guides(colour = guide_legend(override.aes = list(size = 1, alpha = 1)))+ #override the transparency and size to make legend readable
  scale_color_viridis(discrete = T, begin = 0.3) #add colour theme

map_pp

```

#Analyse data
```{r}
library(tidyverse) #easy manipulation
library(lme4) #mixed models
library(lmerTest) #p-values in summary()

#lm_basic <- lmer(doy ~ sensitivity_yn*year + (1|species), data = df) #original model with species and binary sensitivity

lm_binary <- lmer(doy ~ sensitivity_yn*year-1 + (1|species) + lat:sensitivity_yn, data = df) #improved model with latitude as fixed effect and nicer looking intercepts

lm_final_init <- lmer(doy ~ simple_photoperiodism*year-1 + (1|species) + lat, data = df)#improved model with photoperiod groups instead of binary but single latitude effect

lm_comparison <- lmer(doy ~ year + (1|species) + lat:simple_photoperiodism, data = df) #comparison model to lm_final with photoperiodism removed as variable

lm_final <- lmer(doy ~ simple_photoperiodism*year-1 + (1|species) + lat:simple_photoperiodism, data = df)#improved model with seperate latitude effects

summary(lm_final)
summary(lm_binary)
```

#check assumptions of model
```{r}
qqnorm(resid(lm_final)) #qqplot
qqline(resid(lm_final)) #add line

hist(resid(lm_final), breaks = 25) #look at normality of residuals
```
The qq plot seems a bit heavy tailed at the start but after looking at the histogram, the proportion of samples in the tail is very small compared to the main section, which fits pretty well. 

#Comparing our final model with the others
To determine whether photoperiod information improves our flowering time model, let's see compare the pair
```{r}
AIC(lm_final) #compare the pair
AIC(lm_comparison)

anova(lm_final, lm_comparison) #compare the pair statistically
```
The AICs on their own suggest that photoperiod does improve the model, and the ANOVA shows that this improvement is significant!

#Comparing our final models through bootstrapping
```{r}
library(pbkrtest)

bootstrap <- PBmodcomp(lm_final, lm_comparison, nsim = 5999) #bootstrap comparison
bootstrap
```
Yes they are significantly different

So now we've confirmed that photoperiodism has an effect on flowering time. 
Let's look at the details

#Look at what our model tells us
```{r}
summary(lm_final) #look at our model
```
We have a lot of variable to unpack, so remembering that day neutral is the baseline for when sd or ld is mentioned:

- base flowering time advancement is -0.09 days/year
- ld plants have a modifier of +0.11 days/year, so they move +0.2 days/year
- sd plants have no significant modifier, so they move the same as the base -0.09 days/year

- latitude has an effect of +1.1 days/degree for ld and na plants
- but a very different -3.7 days/degree for short day plants!

#look at each photoperiod group seperately and in emtrends to check the main results
```{r}
lm_na <- lmer(doy ~ year + (1|species) + lat, data = df, subset = simple_photoperiodism == "na")

lm_sd <- lmer(doy ~ year + (1|species) + lat, data = df, subset = simple_photoperiodism == "sd")

lm_ld <- lmer(doy ~ year + (1|species) + lat, data = df, subset = simple_photoperiodism == "ld")

#DAY NEUTRAL
summary(lm_na)
#SHORT DAY
summary(lm_sd)
#LONG DAY
summary(lm_ld)

```
#checking the differences between the groups

```{r}
library(emmeans)
emtrends(lm_final, pairwise~simple_photoperiodism, var = "year")
```

These trends seem to be pretty consistent with our initial results.

#plot graphs
```{r}
library(viridis) #load a nice and colourblind-friendly palette

mypalette = viridis(12) #viridis again

ggplot.photogroups <- 
  ggplot(subset(df))+
  geom_point(aes(y = doy, x = year, colour = simple_photoperiodism), size = 1, alpha = 0.2)+
  xlab("Year")+
  ylim(-4, 366)+
  xlim(1950, 2017)+
  ylab("Average First Flowering Day")+
  guides(color = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE)) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"))

ggplot.species <- ggplot(subset(df))+
  geom_smooth(aes(y = doy, x = year, group = species, colour = simple_photoperiodism), method = "lm", se = F)+
  xlab("Year")+
  ylim(-4, 366)+
  xlim(1950, 2017)+
  ylab("Average First Flowering Day")+
  scale_color_viridis(discrete = T) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"))
```

#graph each simple photogroup with the following details:
- trendline info from individual lm models
- trendline assumes mean latitude for each group

#best to print 8*6
```{r}
ggplot.photogroups.main <- 
  ggplot(df)+
  geom_point(aes(y = doy, x = year, 
                 colour = simple_photoperiodism, 
                 shape = simple_photoperiodism), 
             size = 1, alpha = 0.5)+
  scale_shape_manual(name = "Photoperiod Sensitivity", 
                     values = c(3, 1, 2), # o,+,x
                     labels = c("Day Neutral", "Long Day","Short Day"))+
  scale_colour_manual(name = "Photoperiod Sensitivity", 
                      values = c(mypalette[3], mypalette[8], mypalette[11]),
                      labels = c("Day Neutral", "Long Day", "Short Day")) +
  geom_segment(aes(x = 1950, y = 135.3, xend = 2017, yend = 129.0), colour = mypalette[3], size = 1)+ #na
    geom_segment(aes(x = 1950, y = 146.9, xend = 2017, yend = 148.1), colour = mypalette[5], size = 1)+ #ld
    geom_segment(aes(x = 1950, y = 172.1, xend = 2017, yend = 162.5), colour = mypalette[10], size = 1)+ #sd 
  xlab("Year")+
  ylim(-1, 366)+
  xlim(1949, 2017)+
  ylab("Average First Flowering Day")+
  guides(color = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE), shape = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE)) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"),
        plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm"))

```

#a graph with no transparency in the points to pull the legend from
```{r}
#no other
ggplot.photogroups.legend <- 
  ggplot()+
  geom_point(aes(y = doy, x = year, colour = simple_photoperiodism), size = 3, alpha = 1)+
  scale_colour_manual(values = c(mypalette[3], mypalette[8], mypalette[11]), labels = c("Day Neutral", "Long Day","Short Day"))+
  xlab("Year")+
  ylim(-4, 366)+
  xlim(1950, 2017)+
  ylab("Average First Flowering Day")+
  guides(color = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE)) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"),
        plot.margin = margin(1, 1, 1, 1, "cm"))

```

```{r}
#average latitude of all points
mean(df$lat)

#average latitude grouped by photoperiod binary
df %>% group_by(sensitivity_yn)%>% summarize(mean = mean(lat, na.rm = TRUE))

#average latitude grouped by photoperiod sensitivity
df %>% group_by(simple_photoperiodism)%>% summarize(mean = mean(lat, na.rm = TRUE))
```


