---
title: "R Notebook"
output: html_notebook
---

This code analyses whether woodiness or life history is correlated with the different types of photoeperiod sensitivity cues in flowering. 

For more information, please contact:
karen.zeng@unsw.edu.au

##load packages
```{r}
library(tidyverse)
#library(ggplot2) #for graphs
#library(viridis) #for colours

#library(viridis)
#mypalette = viridis(12) #viridis over 12 segments, 1=yellow 4=green 8=blue 12=purple
```

##load photoperiod sensitivity data and filter out complex photoperiod categories
```{r}
pp <- read.csv("./data/appendix1.csv") #read in photoperiod data
pp <- pp %>% filter(photoperiodism == "sd" |photoperiodism == "ld" | photoperiodism == "na")
```

##load woodiness data from [GlobalWoodinessDatabase]
```{r}
wood <- read.csv("./data/GlobalWoodinessDatabase.csv")

wood <- transmute(wood, species = gs, woodiness = woodiness)

wood$woodiness[wood$woodiness == "H"] <- "herbaceous"
wood$woodiness[wood$woodiness == "W"] <- "woody"

table(wood$woodiness)
```

##load longevity data from [TRY]
```{r}
longevity <- read.csv(file = "./data/TRYlongevity.csv")

longevity <- longevity %>% 
  mutate(ann_per = case_when(OrigValueStr == "annual" | 
                               OrigValueStr == "Annual" | 
                               OrigValueStr == "annuals" |
                               OrigValueStr == "summer annuals" |
                               OrigValueStr == "always annual" |
                               OrigValueStr == "winter annuals" |
                               OrigValueStr == "annual-winter annual" |
                               OrigValueStr == "winter annual" |
                               (OriglName == "Life history" & OrigValueStr == "1" ) |
                               (OriglName == "Plant phenology: Annual" & OrigValueStr == "yes" )     ~   "annual",  ######annuals
                             OrigValueStr == "perennial" | 
                               OrigValueStr == "Perennial" | 
                               OrigValueStr == "perennials" |                                   
                               OrigValueStr == "always pluriennial-pollakanthic" | 
                               (OriglName == "Plant phenology: Biennial" & OrigValueStr == "yes" ) | 
                               OrigValueStr == "perennial < 20 years" | 
                               OrigValueStr == "woody" | 
                               OrigValueStr == "perennial/woody" | 
                               OrigValueStr == "perennial > 20 years" | 
                               OrigValueStr == "poly-annuals > 50 years (long-lived perennials)" | 
                               OrigValueStr == "always biennial, always pluriennial-hapaxanthic" | 
                               OrigValueStr == "always biennial, always pluriennial-pollakanthic" | 
                               OrigValueStr == "tree" | 
                               OrigValueStr == "shrub" | 
                               OrigValueStr == "always pluriennial-hapaxanthic, always pluriennial-pollakanthic" | 
                               OrigValueStr == "always pluriennial-hapaxanthic" | 
                               OrigValueStr == "biennial" | 
                               OrigValueStr == "annual/biennial" | 
                               OrigValueStr == "poly-annuals < 5 years (short-lived perennials)" | 
                               OrigValueStr == "Biennial" | 
                               OrigValueStr == "biennial/perennial" | 
                               OrigValueStr == "always biennial" | 
                               OrigValueStr == "biennial-perennial" | 
                               OrigValueStr == "sometimes biennial, always pluriennial-hapaxanthic, sometimes pluriennial-pollakanthic" | 
                               OrigValueStr == "sometimes biennial, sometimes pluriennial-hapaxanthic, always pluriennial-pollakanthic" | 
                               OrigValueStr == "biennial/perennial/woody" | 
                               OrigValueStr == "sometimes biennial, always pluriennial-pollakanthic" | 
                               OrigValueStr == "poly-annuals 5-50 years (medium-lived perennials)" | 
                               (OriglName == "Plant phenology: Perennial" & OrigValueStr == "yes" )| 
                               (OriglName == "Plant phenology: Annual" & OrigValueStr == "no" )    ~   "perennial"  ###perennials
  ))

longevity <- transmute(longevity, species = SpeciesName, life_history = ann_per) #isolate info

longevity <- na.omit(longevity) #remove NAs

longevity <- longevity[!duplicated(longevity$species), ] #remove duplicates

table(longevity$life_history)
```

##combine the datasets
```{r}
pp <- left_join(pp, wood, by = "species") #attach woodiness dataset to main 
pp <- left_join(pp, longevity, by = "species") #attach longevity dataset to main

rm(longevity) #remove both datasets from environment
rm(wood)
```

#Life History Analysis
```{r}
longevity <- pp %>% #Tally life histories
  filter(!is.na(life_history))%>% #remove records without life history
  filter(photoperiodism != "other") %>% #remove records with unclear photoperiodism
  group_by(photoperiodism, life_history) %>%
  tally()

longevity_chisq <- chisq.test(longevity$n) #Perform chi-square test

perennial_pie <- longevity %>% #choose our dataframe
  filter(life_history == "perennial") %>% #Filter so we are only looking perennials
  ggplot(aes(x = "", y = n, fill = photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Perennial")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= "#FC4E07", "ld"= "#E7B800", "na"= "#00AFBB")) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels

annual_pie <- longevity %>% #choose our dataframe
  filter(life_history == "annual") %>% #Filter so we are only looking at annuals
  ggplot(aes(x = "", y = n, fill = photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Annual")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= "#FC4E07", "ld"= "#E7B800", "na"= "#00AFBB")) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels
```

#Look at life history results
```{r}
longevity_chisq
longevity_chisq$residuals
longevity_chisq$expected
longevity_chisq$observed

perennial_pie
annual_pie

```

Annual species seem to be more likely to be photoperiod-sensitive overall, and seem to be more short day than long day, although they exist in both groups.

#Woodiness analysis
```{r}
woodiness <- pp %>% #Tally woodiness
  filter(!is.na(woodiness))%>% #remove records without woodiness
  filter(woodiness != "variable") %>% #remove records with unclear woodiness
  filter(photoperiodism != "other") %>% #remove records with unclear photoperiodism
  group_by(photoperiodism, woodiness) %>%
  tally()

woodiness_chisq <- chisq.test(woodiness$n) #Perform chi-square test

woody_pie <- woodiness %>% #choose our dataframe
  filter(woodiness == "woody") %>% #Filter so we are only looking at herbaceous species
  ggplot(aes(x = "", y = n, fill = photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Woody")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= "#FC4E07", "ld"= "#E7B800", "na"= "#00AFBB")) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels


herbaceous_pie <- woodiness %>% #choose our dataframe
  filter(woodiness == "herbaceous") %>% #Filter so we are only looking at this type of plant
  ggplot(aes(x = "", y = n, fill = photoperiodism)) + #single graph split by proportion of photoperiod sensitivity
  ggtitle("Herbaceous")+ #add title
  geom_bar(width = 1, stat = "identity") + #make it a bar graph
  theme_minimal() + #removes most of the background fluff
  theme( #removes the rest of the background fluff
    axis.title.x = element_blank(),
    axis.text.x=element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
    ) +
  coord_polar("y", start=0)+ #turns bar chart into a pie
  scale_fill_manual(values = c("sd"= "#FC4E07", "ld"= "#E7B800", "na"= "#00AFBB")) + #add colour
  labs(fill = "Photoperiod Sensitivity") #add labels

```

#woodiness results
```{r}
woodiness_chisq
woodiness_chisq$residuals
woodiness_chisq$expected
woodiness_chisq$observed

woody_pie
herbaceous_pie
```
Both herbaceous and woody species are just as likely to be photoperiod sensitive, but it seems that woody species are more likely to be short day, and conversely herbaceous species are more likely to be long-day.


```{r}

```

