---
title: "PhotoperiodNotebook2_Analysis"
output: html_notebook
---

This code analyses whether the different flowering time changes (1950-2017) are correlated with the photoperiod sensitivity cues required for each species to flower. It also generates the df.csv file required for further analyses.

For more information, please contact:
karen.zeng@unsw.edu.au

##load packages
```{r}
library(tidyverse)#general data wrangling
library(ggplot2) #for graphs

library(lme4) #mixed models
library(lmerTest) #p-values in summary()
```

#load viridis palette
```{r}
library(viridis) #load a colourblind-friendly palette
mypalette = viridis(12) #viridis over 12 segments, 1=yellow 4=green 8=blue 12=purple
```

##load data
```{r}
pp <- read.csv("./data/SI1_Photoperiod_photoperiod_data.csv") #read in photoperiod data

pp <- pp %>% 
  group_by(species)%>% #for each species
  slice(which.min(order))  #find most accurate photoperiod record based on recent-ness and number of treatments

ft <- read.csv("./data/SI2_Photoperiod_first_flowering_day_data.csv") #read completed flowering time data

df <- left_join(ft, pp, by = "species") #join the two datasets
rm(pp)
rm(ft)
```

##Define simplified photoperiodism field to group all the different types of photoperiod cues
```{r}
df$simple_photoperiodism <- "other" #if something is not ld, sd or na it will be grouped as 'other'

df$simple_photoperiodism[df$photoperiodism == "na"] <-"na"
df$simple_photoperiodism[df$photoperiodism == "sd"] <-"sd"
df$simple_photoperiodism[df$photoperiodism == "ld"] <-"ld"

#write.csv(df, "./data/df.csv")
```

##Remove borderline cases of photoperiodism from analysis
```{r}
df <- df %>% subset(simple_photoperiodism != "other")#only keep ld, sd and na data

df$simple_photoperiodism <- as.factor(df$simple_photoperiodism) #make it a factor
df$simple_photoperiodism <- relevel(df$simple_photoperiodism, "na") #relevel so na is the reference factor

table(df$simple_photoperiodism)

```

#General look at combined data
```{r}
unique(df$photoperiodism) #print all the different types of photoperiod cues found

unique(df$species) #print all the different species covered by both datasets

plot(df$year, df$doy) #plot of whenever something first flowered throughout the years (very overlapped)
```

## plotting our data geographically
```{r}
summary(df)

#You only need these packages if you want to map the data

library(sf) #for applying graphs to spatial data (simple version of sp)
library(rgeos) #for gCentroid()
library(rworldmap) #for getMap()
library(rworldxtra) #for higher resolution map

map_base <- ggplot() + #specify map base
  borders("world", colour="gray50", fill="gray50")

map_pp <- map_base + #use map base
  geom_jitter(data = df, aes(x = lon, y = lat, colour = simple_photoperiodism), size = 0.05, alpha = 0.1, shape = 1, width = 0.4, height = 0.15) +#layer points on map
  xlim(-125, 45)+ #set x limits based on summary
  ylim(20, 65)+#set y limits based on summary
  guides(colour = guide_legend(override.aes = list(size = 1, alpha = 1)))+ #override the transparency and size to make legend readable
  scale_color_viridis(discrete = T, begin = 0.3) #add colour theme

map_pp

```

Unfortunately, our data is not very evenly distributed. Hopefully this issue will be resolved in the future with more citizen science and large scale data initiatives.

```{r}
remove(map_base, map_pp)
```


#Analyse data
```{r}
#I've left some previous iterations of our model commented out so you can see our progress!

#lm_basic <- lmer(doy ~ sensitivity_yn*year + (1|species), data = df) #original model with species and binary sensitivity

#lm_binary <- lmer(doy ~ sensitivity_yn*year + (1|species) + lat:sensitivity_yn, data = df) #improved model with latitude

#lm_final_init <- lmer(doy ~ simple_photoperiodism*year + (1|species) + lat:simple_photoperiodism, data = df)#improved model with photoperiod groups instead of binary

lm_comparison <- lmer(doy ~ year + (1|species) + lat, data = df) #comparison model to lm_final without photoperiod variables

lm_final <- lmer(doy ~ simple_photoperiodism*year + (1|species) + lat:simple_photoperiodism, data = df)#improved model with all changes as above and seperate latitude effects

```

#check assumptions of model
```{r}
qqnorm(resid(lm_final)) #qqplot
qqline(resid(lm_final)) #add line

hist(resid(lm_final), breaks = 100) #look at normality of residuals
```
The qq plot seems a bit heavy tailed at the start but after looking at the histogram, the proportion of samples in the tail is very small compared to the main section, which fits pretty well. 

#Looking at the linear model
```{r}
summary(lm_final)
```

We have a lot of variable to unpack, so remembering that day neutral is the baseline for when sd or ld is mentioned:

##Day neutral and short day plants are advancing at almost a day per decade, but long day plants are not!
- base flowering time advancement is -0.09 days/year (for plants with no photoperiod requirements)
- ld plants have a significant modifier of +0.11 days/year, so they move +0.2 days/year
- sd plants have no significant difference to day neutral, so they move the same as the base -0.09 days/year 

##As you move away from the equator, short day and day neutral plants flower later. But short day plants flower earlier.
- latitude has an effect of +1.1 days/degree for ld and na plants
- but a very different -3.7 days/degree for short day plants!

#Comparing our final model with the others
To determine whether photoperiod information improves our flowering time model, let's see compare the pair
```{r}
aic_final <- AIC(lm_final) #compare the pair
aic_comparison <- AIC(lm_comparison)

anova_comparison_final <- anova(lm_comparison, lm_final) #compare the pair statistically

Anova(lm_final) #also check if photoperiodism has an effect on the interaction between photoperiod and year
```
The AICs on their own suggest that photoperiod does improve the model, and the ANOVA shows that this improvement is significant!

#Optional: Comparing our final models through bootstrapping
```{r}
#library(pbkrtest)

#bootstrap <- PBmodcomp(lm_final, lm_comparison, nsim = 5999) #bootstrap comparison
```
Yes they are significantly different according to this method too :D

#look at each photoperiod group seperately and in emtrends to check the main results
```{r}
lm_na <- lmer(doy ~ year + (1|species) + lat, data = df, subset = simple_photoperiodism == "na")

lm_sd <- lmer(doy ~ year + (1|species) + lat, data = df, subset = simple_photoperiodism == "sd")

lm_ld <- lmer(doy ~ year + (1|species) + lat, data = df, subset = simple_photoperiodism == "ld")

#DAY NEUTRAL
summary(lm_na)
#SHORT DAY
summary(lm_sd)
#LONG DAY
summary(lm_ld)

```
#checking the differences between the groups

```{r}
library(emmeans)
em <- emtrends(lm_final, pairwise~simple_photoperiodism, var = "year")
em
```

These trends seem to be pretty consistent with our initial results. 

Short day plants are similar to Day Neutral (na) plants, while Long day plants are advancing in the opposite direction, if at all.

#plot graphs
Unfortunately, it get a bit tricky to plot a graph that fully represents the results we get when using random effects, but will try our best.

```{r}
ggplot.photogroups <- 
  ggplot(subset(df))+
  geom_point(aes(y = doy, x = year, colour = simple_photoperiodism), size = 1, alpha = 0.2)+
  xlab("Year")+
  ylim(-4, 366)+
  xlim(1950, 2017)+
  ylab("Average First Flowering Day")+
  guides(color = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE)) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"))

ggplot.species <- ggplot(subset(df))+
  geom_smooth(aes(y = doy, x = year, group = species, colour = simple_photoperiodism), method = "lm", se = F)+
  xlab("Year")+
  ylim(-4, 366)+
  xlim(1950, 2017)+
  ylab("Average First Flowering Day")+
  scale_color_viridis(discrete = T) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"))
```

#graph each simple photogroup with the following details:
- trendline info from individual lm models
- trendline assumes mean latitude for each group

#best to print 8*6
```{r}
ggplot.photogroups.main <- 
  ggplot(df)+
  geom_point(aes(y = doy, x = year, 
                 colour = simple_photoperiodism, 
                 shape = simple_photoperiodism), 
             size = 1, alpha = 0.5)+
  scale_shape_manual(name = "Photoperiod Sensitivity", 
                     values = c(3, 1, 2), # o,+,x
                     labels = c("Day Neutral", "Long Day","Short Day"))+
  scale_colour_manual(name = "Photoperiod Sensitivity", 
                      values = c(mypalette[3], mypalette[8], mypalette[11]),
                      labels = c("Day Neutral", "Long Day", "Short Day")) +
  geom_segment(aes(x = 1950, y = 135.3, xend = 2017, yend = 129.0), colour = mypalette[3], size = 1)+ #na
    geom_segment(aes(x = 1950, y = 146.9, xend = 2017, yend = 148.1), colour = mypalette[5], size = 1)+ #ld
    geom_segment(aes(x = 1950, y = 172.1, xend = 2017, yend = 162.5), colour = mypalette[10], size = 1)+ #sd 
  xlab("Year")+
  ylim(-1, 366)+
  xlim(1949, 2017)+
  ylab("Average First Flowering Day")+
  guides(color = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE), shape = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE)) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"),
        plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm"))
```

##second version of graph for nicer looking publication.
```{r}
ggplot.photogroups.main2 <- 
  ggplot(df)+
  geom_point(aes(y = doy, x = year, 
                 colour = simple_photoperiodism, 
                 shape = simple_photoperiodism), 
             size = 1, alpha = 0.5)+
  scale_shape_manual(name = "Photoperiod Sensitivity", 
                     values = c(0, 1, 2), # o,+,x
                     labels = c("Day Neutral", "Long Day","Short Day"))+
  scale_colour_manual(name = "Photoperiod Sensitivity", 
                      values = c("#00AFBB", "#E7B800", "#FC4E07"),
                      labels = c("Day Neutral", "Long Day", "Short Day")) +
  geom_segment(aes(x = 1950, y = 135.3, xend = 2017, yend = 129.0), colour = "#00AFBB", size = 1)+ #na
    geom_segment(aes(x = 1950, y = 146.9, xend = 2017, yend = 148.1), colour = "#E7B800", size = 1)+ #ld
    geom_segment(aes(x = 1950, y = 172.1, xend = 2017, yend = 162.5), colour = "##FC4E07", size = 1)+ #sd 
  xlab("Year")+
  ylim(-1, 366)+
  xlim(1949, 2017)+
  ylab("Average First Flowering Day")+
  guides(color = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE), shape = guide_legend(title = "Photoperiod Sensitivity", reverse = TRUE)) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15, face  = "bold"),
        plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm"))
```

```{r}
ggplot.photogroups.main
```


```{r}
ggplot.photogroups.main2
```

