---
title: "R Notebook"
output: html_notebook
---
#load packages
```{r}
library(tidyverse)
library(ggplot2) #for graphs
library(ggmap) #for mapping
library(sf) #for applying graphs to spatial data (simple version of sp)
library(raster) #for raster format
library(rgeos) #for gCentroid()
library(rworldmap) #for getMap()
library(rworldxtra) #for higher resolution map
```
#load data
Mypalette:
- a colour palette to make the graph colours consistent

Photoperiod data:
- species
- photoperiodism: photoperiod sensitivity requirements(ad, id, ld, sd, ldsd, lsd, sd, sld, yes (exists in unknown form), na)
- cultivar: if present, cultivar name
- reference: source of information
- notes: additional notes on photoperiod sensitivity
- sensitivity_yn: binary variable of whether photoperiod sensitivity in any form exists in species

Flowering time data:
- species
- year: year of record
- doy: day of year
- source: source of information

```{r}
pp <- read.csv("./Data/appendix1.csv") #read in photoperiod data
pp$species <- gsub("_", " ",pp$species)
ft <- read.csv("./data/appendix2.csv") #read completed flowering time data

df <- left_join(ft, pp, by = "species")
rm(pp)
rm(ft)

```

#simple map of data
```{r}
map_base <- ggplot() + #specify map base
  borders("world", colour="gray50", fill="gray50")

map_pp_reference <- map_base + #use map base
  geom_jitter(data = df, aes(x = lon, y = lat, colour = reference), size = 0.05, alpha = 0.1, shape = 1) +#layer points on map
  guides(colour = guide_legend(override.aes = list(size = 1, alpha = 1))) #override the transparancy and size to make legend readable

map_pp_reference

```

#create raster
```{r}

```

